---
title: "OmicsLens: End-to-End Multi-Omics Integration Workflow"
author: "Priyansh Pathak"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{OmicsLens workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8, fig.height = 5,
  message  = FALSE, warning = FALSE
)
```

# Introduction

**OmicsLens** provides a five-function pipeline for integrating RNA-Seq, whole-
genome sequencing (WGS), and DNA methylation data.
The core engine is **MOFA2** (Multi-Omics Factor Analysis v2), which learns
shared latent factors across all three data types. Downstream analyses —
differential expression, gene set enrichment, DMR calling, and survival — are
all driven by those factors, giving results that reflect the *joint* biology
rather than layer-by-layer snapshots.

```
omicslens_load()         ── read + align
       ↓
omicslens_preprocess()   ── normalise + filter
       ↓
omicslens_integrate()    ── MOFA2 latent factors
       ↓
omicslens_analyze()      ── DE · GSEA · DMR · survival
       ↓
omicslens_app()          ── Shiny dashboard
omicslens_report()       ── HTML report
```

# Installation

```r
# Install Bioconductor dependencies first
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("MOFA2", "DESeq2", "fgsea", "DMRcate"))

# Install OmicsLens from GitHub
# install.packages("devtools")
devtools::install_github("priyanshpathak/OmicsLens")
```

# Step 1 — Load data

`omicslens_load()` accepts file paths (CSV / TSV / RDS) or in-memory R objects.
Only one layer is required; the others can be added later.

```{r load-data}
library(OmicsLens)
set.seed(42)

# ── Simulate three data layers ────────────────────────────────────────────────
n_samples <- 40L
n_genes   <- 500L
n_cpgs    <- 2000L
n_mut     <- 150L

sample_ids <- paste0("Sample_", seq_len(n_samples))

rna_counts <- matrix(
  rpois(n_genes * n_samples, lambda = 80),
  nrow     = n_genes,
  dimnames = list(paste0("Gene",  seq_len(n_genes)),  sample_ids)
)

meth_beta <- matrix(
  rbeta(n_cpgs * n_samples, 2, 5),
  nrow     = n_cpgs,
  dimnames = list(paste0("cg", seq_len(n_cpgs)), sample_ids)
)

var_binary <- matrix(
  rbinom(n_samples * n_mut, 1, 0.08),
  nrow     = n_samples,
  dimnames = list(sample_ids, paste0("MutGene", seq_len(n_mut)))
)

metadata <- data.frame(
  sample_id = sample_ids,
  condition = rep(c("Tumour", "Normal"), n_samples / 2),
  time_os   = runif(n_samples, 6, 60),
  event     = rbinom(n_samples, 1, 0.5),
  row.names = sample_ids,
  stringsAsFactors = FALSE
)

# ── Load ──────────────────────────────────────────────────────────────────────
obj <- omicslens_load(
  rna_counts  = rna_counts,
  methylation = meth_beta,
  variants    = var_binary,
  metadata    = metadata,
  variant_format = "binary_matrix"
)
print(obj)
```

# Step 2 — Preprocess

Each layer is normalised independently:

| Layer       | Method                                               |
|-------------|------------------------------------------------------|
| RNA-Seq     | Low-count gene removal → DESeq2 VST normalisation   |
| Variants    | Filter genes by mutation allele frequency            |
| Methylation | NA filtering → beta→M-value conversion → top-N CpGs |

```{r preprocess}
obj <- omicslens_preprocess(
  obj,
  rna_min_counts   = 10,
  variant_min_maf  = 0.05,
  meth_var_top_n   = 1000
)
```

# Step 3 — MOFA2 Integration

```{r integrate, eval=FALSE}
# Requires Python + mofapy2 (managed via basilisk automatically)
obj <- omicslens_integrate(obj, n_factors = 10, seed = 42)
```

```{r integrate-fake, echo=FALSE}
# Synthetic factor scores for illustration (skips Python requirement in vignette)
set.seed(42)
factors_mat <- matrix(rnorm(n_samples * 10), nrow = n_samples,
                       dimnames = list(sample_ids, paste0("Factor", 1:10)))
obj$results$mofa <- list(
  factors = factors_mat,
  weights = lapply(c("RNA","Variants","Methylation"), function(v) {
    list(matrix(rnorm(n_genes * 10), n_genes, 10,
                 dimnames = list(paste0("Gene", seq_len(n_genes)),
                                 paste0("Factor", 1:10))))
  }) |> setNames(c("RNA","Variants","Methylation")),
  variance_explained = list(
    r2_per_factor = list(
      data.frame(RNA = runif(10, 0, 0.1),
                 Variants = runif(10, 0, 0.05),
                 Methylation = runif(10, 0, 0.08))
    )
  )
)
# Fake MOFA model indicator
obj$mofa_model <- TRUE
```

After integration, factor scores and weights per view are available:

```{r factor-head}
head(obj$results$mofa$factors[, 1:4])
```

## Plot: factor scatter

```{r plot-factors}
omicslens_plot_factors(obj, factor_x = 1, factor_y = 2)
```

# Step 4 — Downstream Analysis

```{r analyze, eval=FALSE}
obj <- omicslens_analyze(
  obj,
  run_de             = TRUE,
  run_gsea           = TRUE,
  run_dmr            = TRUE,
  run_survival       = TRUE,
  survival_time_col  = "time_os",
  survival_event_col = "event"
)
```

```{r analyze-fake, echo=FALSE}
# Synthetic DE results for illustration
set.seed(1)
n_de <- nrow(rna_counts)
de_df <- data.frame(
  gene           = rownames(rna_counts),
  baseMean       = abs(rnorm(n_de, 100, 50)),
  log2FoldChange = rnorm(n_de),
  lfcSE          = abs(rnorm(n_de, 0.3, 0.1)),
  stat           = rnorm(n_de),
  pvalue         = runif(n_de),
  padj           = p.adjust(runif(n_de), method = "BH"),
  stringsAsFactors = FALSE
)
de_df <- de_df[order(de_df$padj), ]
obj$results$de <- de_df

# Synthetic GSEA results
obj$results$gsea <- data.frame(
  pathway = c("HALLMARK_E2F_TARGETS","HALLMARK_MYC_TARGETS_V1",
              "HALLMARK_INFLAMMATORY","HALLMARK_APOPTOSIS"),
  NES     = c(2.1, 1.8, -1.6, -2.0),
  pval    = c(0.001, 0.01, 0.02, 0.005),
  padj    = c(0.005, 0.04, 0.08, 0.02),
  size    = c(200, 150, 180, 120),
  stringsAsFactors = FALSE
)

# Groups
groups <- ifelse(factors_mat[, 1] >= median(factors_mat[, 1]), "High", "Low")
names(groups) <- sample_ids
obj$results$groups <- groups
```

## Volcano plot

```{r volcano}
omicslens_plot_de(obj, padj_cutoff = 0.05, lfc_cutoff = 1)
```

## GSEA dot plot

```{r gsea}
omicslens_plot_gsea(obj)
```

## Survival

```{r survival, eval=FALSE}
omicslens_plot_survival(obj)
```

# Step 5 — Shiny Dashboard

```{r shiny, eval=FALSE}
omicslens_app(obj)
```

The dashboard provides seven tabs covering every analysis result, with
interactive plotly charts, DT tables, and download buttons.

# Step 6 — HTML Report

```{r report, eval=FALSE}
omicslens_report(
  obj,
  output_file = "my_analysis.html",
  title       = "TCGA-BRCA Multi-Omics",
  author      = "Priyansh Pathak"
)
```

# Session Info

```{r session}
sessionInfo()
```
